---
title: "phosphoProcessR: Differential Phosphorylation Analysis (DPA)"
author: "Ashley J. Waardenberg"
date: 'Last modified: 2019-05-16. Compiled: `r Sys.Date()`'
output:
  html_document:
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document:
    toc: yes
    toc_depth: '3'
references:
- DOI: XX
  URL: XX
  author:
  - family: Engholm-Keller
    given: K
  - family: Waardenberg
    given: AJ
  - family: et
    given: al
  container-title: XX
  id: Kasper2018
  issue: X
  issued:
    month: X
    year: 2018
  page: XX-XX
  publisher: XX
  title: in press
  type: article-journal
  volume: XX
- DOI: 10.1007/978-1-4939-6955-5_17
  URL: https://www.ncbi.nlm.nih.gov/pubmed/28477123
  author:
  - family: Waardenberg
    given: AJ
  container-title: XX
  id: Waardenberg2017
  issued:
    month: May
    year: 2017
  page: 229-244
  publisher: Methods in Molecular Biology
  title: Statistical analysis of ATM-dependent signaling in quantitative mass spectrometry phosphoproteomics
  type: article-journal
  volume: 1599
vignette: >
  %\VignetteIndexEntry{phosphoProcessR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction to phosphoProcessR

phosphoProcessR provides core utilities for performing Differential Phosphorylation Analysis (DPA). It has functionality for remapping and annotating an "evidence.txt"" file from a MaxQuant output, given a protein database, ready for further statistical analysis. Functionality includes normalisation, missing value imputation, batch correction utilising SVA and statistical analysis of differential phosphorylation using limma. Currently it only supports pairwise analysis between two groups and TMT labelled samples, however will be expanded to include additional normalisation, imputation, correction and differential analysis methods. phosphoProcessR implements an updated version described in: @Waardenberg2017 and @Kasper2018.

phosphoProcessR contains 3 core functions:

+ **```tidyEvidence```** remap and merge MaxQuant Evidence file into "tidy" data frame
+ **```phosphoDE```** perform Differential Phosphorylation Analysis (DPA)
+ **```plot_this```** function which includes wrappers of various plotting functions

Detailed information for each of these functions can be accessed using the ```?``` command before the function of interest. E.g. ```?tidyEvidence```

# phosphoProcessR example workflow

We will now consider an example dataset to format an "evidence.txt" file output by MaxQuant and perform DPA. These example data are provided in the phosphoProcessR package and relate to a dataset containing post-translational modifications dervied from Plasmodium falciparum-infectedhuman infected and non-infected human placenta's. See original publication for more details: Kawahara et al. MCP 2018 \url{https://doi.org/10.1074/mcp.RA118.000907}.

Begin by loading the phosphoProcessR library and the data libraries included in the package.
```{r echo=TRUE}
library(phosphoProcessR)
data(human_fasta_example)
data(malaria_evidence_example)
data(malaria_annotation_example)

## sample 100 data points for demonstration
evidence_file_example <- head(malaria_evidence_example, 100)
#evidence_file_example <- malaria_evidence_example

# calling the function reannotate and tidy the evidence file for statistical analysis
# tidyEvidence supports multi-core processing using the BiocParallel package:
# load BiocParallel library
library(BiocParallel)
# set/register the number of cores to use
register(SnowParam(workers = 4))

# call function
evidence_tidy <- tidyEvidence(evidence_file = evidence_file_example,
                              annotation_file = malaria_annotation_example,
                              fasta_file = human_fasta_example,
                              window_size = 15,
                              min_prob = 0.75,
                              filter_site_method = "peptide",
                              verbose = TRUE)

# view the data extracted
head(evidence_tidy$intensity, 5)

```

## Plotting Phosphorylation Site Probability distribution

The ouputs of evidence_tidy, using ```tidyEvidence()``` also include data of the phosphosite probabilities. This can be used to inform on the ```min_prob``` threshold utilised. To visualise:

```{r echo=TRUE}
plot_this(data_in = evidence_tidy$site_probability,
          title = "Malaria example",
          density = TRUE)

```

# Perform Differential Phosphorylation Analysis (DPA)

```{r echo=TRUE}

# utilising the intensity values derived from tidyEvidence, we now can perform DPA.
# DPA is currently limited to pairs of groups defined in an annotation_file.

DPA_out <- phosphoDE(phospho_input = evidence_tidy$intensity,
                     annotation_file = malaria_annotation_example,
                     correct = TRUE,
                     adjust_method = "none")


```

## Principle Component Analysis
```{r echo=TRUE}

# get the students to check the results on each of the outputs as an exercise:
plot_this(data_in = DPA_out$corrected_data,
          annotation_file = malaria_annotation_example,
          legend = TRUE,
          label = TRUE,
          title = "Malaria example",
          pca = TRUE,
          hclust = FALSE,
          density = FALSE)
```

## Hierarchicical clustering

```{r echo=TRUE}
plot_this(data_in = DPA_out$corrected_data,
          annotation_file = malaria_annotation_example,
          legend = TRUE,
          label = TRUE,
          title = "Malaria example",
          hclust = TRUE)
```

# References
